On Optimality of Finite Element Integration

### 2015-09-23 TOMS_Paper/assembly.tex
Rewrite bits of section 3

\subsection{Operation Count Minimization}
We here discuss how we can systematically reach the goal set by Definition~\ref{def:optimality}.

We reinforce that eliminating sharing from the outer-product loop nest does not suffice. As suggested in Section~\ref{sec:optimal-impl}, we wonder whether, and under what conditions, the reduction imposed by $L_i$ could be pre-evaluated, thus reducing the operation count.

To partly answer this question, we make use of a result -- the foundation of tensor contraction mode -- from~\citeN{Kirby:TC}. Essentially, multilinear forms can be seen as sums of monomials, each monomial being an integral over the equation domain of products (of derivatives) of functions from discrete spaces; such monomials can always be reduced to a product of two tensors (see Section~\ref{sec:background}). We interpret this result at the loop nest level: with an input as in Figure~\ref{code:loopnest}, we can always dissect \texttt{F} into distinct sub-expressions (the monomials). Each sub-expression is factorizable so as to split constants from $[L_i, L_j, L_k]$-dependent terms. These $[L_i, L_j, L_k]$-dependent terms can be hoisted outside of $\Lambda$, and finally pre-evaluated into temporaries. As part of this process, the reduction induced by $L_i$ is computed and, therefore, disappears from $\Lambda$. This is actually what we were referring to when we introduced the pre-evaluation transformation in Section~\ref{sec:optimal-impl}.

There are two issues: understanding for which monomials pre-evaluation is profitable; coordinating pre-evaluation with sharing elimination, since the latter prevents the former. We start with the second issue. For this, it is convenient to introduce the transformation algorithm. Figure~\ref{code:intuition} provides the pseudo-code of the algorithm. An explanation and a discussion on its optimality follow.

\begin{figure}[h]
\begin{lstlisting}[numbers=left, basicstyle=\small\ttfamily, frame=single]
dissect the input expression into monomials
for each monomial M:
  $\theta_{w}$ = estimate operation count with pre-evaluation
  $\theta_{wo}$ = estimate operation count without pre-evaluation
  if $\theta_{w}$ < $\theta_{wo}$ and memory constraints satisfied:
    mark M as candidate for pre-evaluation
for each monomial M:
  if M does not share terms with M$'$, an unmarked monomial:
    extract M into a separate loop nest
    apply pre-evaluation to M
for each expression:
  remove sharing    
\end{lstlisting}
\caption{Intuition of the transformation algorithm}
\label{code:intuition}
\end{figure}

We study the impact of pre-evaluation, as number of operations saved or introduced, ``locally''; that is, for each monomial, in isolation. If we estimate that, for a given monomial, pre-evaluation will decrease the operation count, then the corresponding sub-expression is extracted, a sequence of transformation steps -- involving expansion, factorization, code motion -- takes place (details in Section~\ref{sec:codegen}), and the evaluation eventually performed. The result is a set of $n$-dimensional tables (these can be seen as ``slices'' of the reference tensor at the math level), $n$ being the arity of the multilinear form. Identical tables are mapped to the same temporary. Eventually, sharing is removed from the resulting expressions by applying a procedure as described in Proposition~\ref{prop:optimal-mln}. The transformed loop nest is as in Figure~\ref{code:loopnest-opt}.

\begin{figure}\begin{CenteredBox}
\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/loopnest_opt.code}
\end{CenteredBox}\caption{A possible loop nest for the local assembly of a form after applying pre-evaluation to some of the monomials (resulting in sub-expressions \texttt{G'}, \texttt{G''}, ...)}\label{code:loopnest-opt}\end{figure}

Before elaborating on the profitability of pre-evaluation (i.e., how to determine the cost function $\theta$ in Figure~\ref{code:intuition}), we need to discuss under which conditions this approach, based on a ``local analysis'' of monomials, is able to reduce loop nests to optimal form.

\begin{Prop}
\label{prop:optimal-approach}
Consider an integration routine originating from an arbitrary multilinear form. A loop nest $\Lambda$, as previously characterized, evaluates an expression comprising a set of monomials, $M$. Let $P$ be the set of monomials for which pre-evaluation is applied because found profitable, and let $Z$ be the set of non pre-evaluated monomials (i.e., $Z = M \setminus P$). Assume that:
\begin{enumerate}
\item the cost function $\theta$ employed is optimal; that is, it predicts correctly whether pre-evaluation is profitable or not for a monomial
\item pre-evaluating different monomials does not result in identical tables
\item monomials in $P$ do not share terms
\end{enumerate} 
Then, the main algorithm in Figure~\ref{code:intuition} leads to inner optimality under the set of memory constraints $C$ (defined in Section~\ref{sec:mem-const}).
\end{Prop}
\begin{proof}
We first comment on the assumptions. 1) We postpone the discussion of how to build $\theta$ to Section~\ref{sec:op_count}. 2) Identical pre-evaluated tables from distinct monomials could be the result of symmetries in tabulated basis functions. This is however a pathological case that we can harmlessly ignore. 3) In complex forms with several monomials, different pre-evaluation candidates could actually share terms. We abstract from this case, which otherwise would require a ``global'' analysis of the monomials in the form that we believe would not justify the gain (if any) in operation count.

We distinguish two classes of loop nests rooted in $\Lambda$: $[L_e, L_j, L_k]$, for the pre-evaluated monomials in $P$, and $[L_e, L_i, L_j, L_k]$, enclosing the remaining monomials in $Z$. Since they only differ for the presence of $L_i$, we relieve the notation by omitting the shared loops when discussing operation counts. The operation count of what we are proving to be the optimal $\Lambda$ synthesis (inferable from Figure~\ref{code:loopnest-opt}) is $\Lambda_{ops} = \Lambda_{ops_1} + \Lambda_{ops_2} = \sum_{\alpha}^{\# P} p_{\alpha} + I \sum_{\beta}^{\# Z} z_{\beta}$, where $p_{\alpha}$ and $z_{\beta}$ represent the operation count of monomials in $P$ and $Z$, respectively, while $I$ is the iteration space size of $L_i$.

We note that, as explained in Section~\ref{sec:mem-const}, $C$ imposes constraints on hoisting. This narrows the proof to demonstrating the following: A) pre-evaluating any $Z_P: Z_P \subseteq Z$ would increase $\Lambda_{ops}$; B) not pre-evaluating any $P_Z: P_Z \subseteq P$ would increase $\Lambda_{ops}$.

A) We prove that $\Lambda_{ops}' = \Lambda_{ops_1}' + \Lambda_{ops_2}' > \Lambda_{ops}$. It is rather obvious that $\Lambda_{ops_1}' \geq \Lambda_{ops_1}$ (it is equal only if, trivially, $Z_P = \emptyset$). We note that if monomials in $Z_P$ share terms with $\overline{Z} = Z \setminus Z_P$, then we have $\Lambda_{ops_2}' = \Lambda_{ops_2}$, so our statement is true. If, on the other hand, at least one monomial does not share any terms, we obtain $\Lambda_{ops_2}' < \Lambda_{ops_2}$ or, equivalently, $\Lambda_{ops_2}' = \Lambda_{ops_2} - I \cdot\delta$. What we have to show now is that even by exposing more pre-evaluations, $\Lambda_{ops_1}' - \Lambda_{ops_1} =  \gamma \geq I \cdot\delta$ holds. Because of assumption 2), the new pre-evaluations cannot expose further sharing. Therefore, the optimality of the cost function (assumption 1) ensures our claim holds.

B) In absence of sharing, the statement is trivially true since we would have $\Lambda_{ops_2}' > \Lambda_{ops_2}$, being the cost function optimal due to assumption 1). Assumption 3) guarantees there cannot be sharing within $P_Z$, which avoids subtle cases in which pre-evaluation would be sub-optimal due to destroying sharing-removal opportunities. The last case we have to consider is when $p \in P_Z$ shares at least one term with $z \in Z$. This situation cannot actually occur by construction: all candidates for pre-evaluation sharing terms with monomials in $Z$ are ``de-classed'' from $P$ to $Z$ (see Figure~\ref{code:intuition}, line 8). The rationale is that since we would have to pay anyway the presence of $z$ in the innermost loop, adding $p$ to $Z$ would not augment the operation count in our model, so we can safely avoid pre-evaluation.

>>>>

\subsection{Reaching Optimality}
We here discuss how we can systematically reach the goal set by Definition~\ref{def:optimality}.

We reinforce the idea that eliminating sharing from the outer-product loop nest does not suffice. As suggested in Section~\ref{sec:optimal-impl}, we wonder whether, and under what conditions, the reduction imposed by $L_i$ could be pre-evaluated, thus reducing the operation count.

To partly answer this question, we make use of a result -- the foundation of tensor contraction mode -- from~\citeN{Kirby:TC}. Essentially, multilinear forms can be seen as sums of monomials, each monomial being an integral over the equation domain of products (of derivatives) of functions from discrete spaces; such monomials can always be reduced to a product of two tensors (see Section~\ref{sec:background}). We interpret this result at the loop nest level: with an input as in Figure~\ref{code:loopnest}, we can always dissect \texttt{F} into distinct sub-expressions (the monomials). Each sub-expression is factorizable so as to split constants from $[L_i, L_j, L_k]$-dependent terms. These $[L_i, L_j, L_k]$-dependent terms can be hoisted outside of $\Lambda$ and stored into temporaries. As part of this process, the reduction induced by $L_i$ is evaluated. Consequently, $L_i$ disappears from $\Lambda$. This transformation is actually what we were referring to when we introduced pre-evaluation in Section~\ref{sec:optimal-impl}.

Two issues should be addressed: understanding for which monomials pre-evaluation is profitable; coordinating pre-evaluation with sharing elimination, since the latter inhibits the former. In this section, we tackle the second issue, while we deal with the second in Section~\ref{sec:op_count}.

We start with introducing the transformation algorithm. Figure~\ref{code:intuition} provides the pseudo-code of the algorithm. An explanation and a discussion about its optimality follow.

%TODO add this somewhere: Further, we note that because of the memory constraints discussed in the previous section, sharing elimination and pre-evaluation fully describe the transformation space.

\begin{figure}
\begin{lstlisting}[numbers=left, basicstyle=\small\ttfamily, frame=single]
dissect the input expression into monomials
for each monomial M:
  $\theta_{w}$ = estimate operation count with pre-evaluation
  $\theta_{wo}$ = estimate operation count without pre-evaluation
  if $\theta_{w}$ < $\theta_{wo}$ and memory constraints satisfied:
    mark M as candidate for pre-evaluation
for each monomial M:
  if M does not share terms with M$'$, an unmarked monomial:
    extract M into a separate loop nest
    apply pre-evaluation to M
for each expression:
  remove sharing    
\end{lstlisting}
\caption{Intuition of the transformation algorithm}
\label{code:intuition}
\end{figure}

Our choice is to study the impact of pre-evaluation, as number of operations saved or introduced, ``locally''; that is, for each monomial, ``in isolation''. If we estimate that, for a given monomial $M$, pre-evaluation will decrease the operation count, then: the sub-expression implementing $M$ is extracted; the surrounding loop nest is suitably ``cloned''; a sequence of transformation steps -- involving expansion, factorization, and code motion -- takes place (details in Section~\ref{sec:codegen}); and, finally, the hoisted code is pre-evaluated using symbolic execution. The result is a set of $n$-dimensional tables (these can be seen as ``slices'' of the reference tensor at the math level), $n$ being the arity of the multilinear form.

Because of symmetries in basis functions, pre-evaluation could produce identical tables. These are mapped to the same temporary. The transformed expression, therefore, can be characterized by sharing. This is why the last two lines of the algorithm, which goal is to eliminate sharing following a procedure as described in Proposition~\ref{prop:optimal-mln}, are transparently applied to all expressions, regardless of whether pre-evaluation was used. The output of the transformation algorithm is as in Figure~\ref{code:loopnest-opt}, assuming the input is still the usual loop nest in Figure~\ref{code:loopnest}.

\begin{figure}\begin{CenteredBox}
\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/loopnest_opt.code}
\end{CenteredBox}\caption{The loop nest produced by the algorithm for an input as in Figure~\ref{code:loopnest}.}\label{code:loopnest-opt}\end{figure}

We elaborate on the optimality of our approach. First of all, we note that, because of the first memory constraint imposed in the previous section, sharing elimination and pre-evaluation fully describe the transformation space. The search for the optimal is then performed by the algorithm in a suitably constructed space. With the following Proposition, we finally aim to formalize the optimality claim.

\begin{Prop}
\label{prop:optimal-approach}
Consider an arbitrary multilinear form and the corresponding loop nest $\Lambda$ implementing it. The form comprises a set of monomials, $M$. Let $P$ be the set of pre-evaluated monomials, and let $Z$ be the set of non pre-evaluated monomials (i.e., $Z = M \setminus P$). Assume that:
\begin{enumerate}
\item the cost function $\theta$ is optimal; that is, it predicts correctly the profitability of pre-evaluation
\item pre-evaluating different monomials does not result in identical tables
\item there is no sharing of terms among monomials in $P$
\end{enumerate} 
Then, the transformation algorithm produces an optimal $\Lambda'$ as established by Definition~\ref{def:optimality}.
\end{Prop}
\begin{proof}
We first comment on the assumptions. 1) We postpone the construction of the oracle $\theta$ to Section~\ref{sec:op_count}. 2) This is inherently a pathological case, from which we abstract. 3) In complex forms with several monomials, different pre-evaluation candidates could actually share terms. We also abstract from this scenario, which would otherwise require a ``global'' analysis of the monomials in $P$.

The operation count for $\Lambda'$ is $\Lambda'_{ops} = \Lambda'_{ops_1} + \Lambda'_{ops_2} = L N O (\sum_{\alpha}^{\# P} p_{\alpha} + I \sum_{\beta}^{\# Z} z_{\beta}$), where $p_{\alpha}$ and $z_{\beta}$ represent operation counts for monomials in $P$ and $Z$, respectively. $\Lambda'_{ops}$ can easily be evinced from Figure~\ref{code:loopnest-opt}.

We prove our claim by demonstrating the following: A) pre-evaluating any $Z_P: Z_P \subseteq Z$ would result in $\Lambda''_{ops}$ such that $\Lambda''_{ops} > \Lambda'_{ops}$; B) not pre-evaluating any $P_Z: P_Z \subseteq P$ would result in $\Lambda''_{ops}$ such that $\Lambda''_{ops} > \Lambda'_{ops}$.

A) It is rather obvious that $\Lambda''_{ops_1} \geq \Lambda'_{ops_1}$ (it is equal only if, trivially, $Z_P = \emptyset$). We note that if monomials in $Z_P$ share terms with $\overline{Z} = Z \setminus Z_P$, then we have $\Lambda''_{ops_2} = \Lambda'_{ops_2}$, so our statement is true. If, on the other hand, at least one monomial does not share any terms, we obtain $\Lambda''_{ops_2} < \Lambda'_{ops_2}$ or, equivalently, $\Lambda''_{ops_2} = \Lambda'_{ops_2} - I \cdot\delta$. What we have to show now is that even by exposing more pre-evaluations, $\Lambda''_{ops_1} - \Lambda'_{ops_1} =  \gamma \geq I \cdot\delta$ holds. Because of assumption 2), the new pre-evaluations cannot expose further sharing. Therefore, the optimality of the cost function (assumption 1) ensures our claim holds.

B) In absence of sharing, the statement is trivially true since we would have $\Lambda''_{ops_2} > \Lambda'_{ops_2}$, being the cost function optimal due to assumption 1). Assumption 3) guarantees there cannot be sharing within $P_Z$, which avoids subtle cases in which pre-evaluation would be sub-optimal due to destroying sharing-removal opportunities. The last case we have to consider is when $p \in P_Z$ shares at least one term with $z \in Z$. This situation cannot actually occur by construction: all candidates for pre-evaluation sharing terms with monomials in $Z$ are ``de-classed'' from $P$ to $Z$ (see Figure~\ref{code:intuition}, line 8). The rationale is that since we would have to pay anyway the presence of $z$ in the innermost loop, adding $p$ to $Z$ would not augment the operation count in our model, so we can safely avoid pre-evaluation.
